Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-02-08 ThesisJP/Background/background.tex
Started adding references.

\section{General-purpose computing on graphics processing units}\label{sec:gpgpu}
In a recent essay, Herb Sutter, a leading authority on software development, reviews the current state of the software and computer industry\todo{cite}. The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004. The two essays spins around the fact that chip manufactures hit a frequency-wall in the beginning of the 21st century. Until then, all mainstream computer programs where typically running in a single thread on a single core, and a programmer could expect the software to annually gain performance without touching the code. This was the "free-lunch" era as depicted in Fig.\,\ref{fig:jungle}. The increase in processing power came mostly from an ever-growing clock frequency. However, power consumption and heat generation where also growing at the same rate, until the level of cooling required finally became to much in 2004.  From that point, chip manufactures had to pursue a different approach  than just increasing the clock frequency. The immediate solution was to embed several cores in one CPU chip. This insured continuation of Moore's law, since the number of transistors per chip could continue to grow. For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to continue to gain performance. It was a concurrency revolution, and  the free lunch was over.

@@

\begin{figure}
\centering
\subfigure[Threads, blocks and grid]{
	\includegraphics[width=0.4\textwidth]{img/cuda_threads.png}
}
\subfigure[Kernel]{
	\includegraphics[width=0.5\textwidth]{img/kernel.pdf}
}
\caption{Depicts how GPU threads are grouped into blocks and arranged in a grid. One thread runs a copy of a kernel function, in this example an element-wise matrix square operation.}
\label{fig:gpu_grid}
\end{figure}

>>>>

\section{General-purpose computing on graphics processing units}\label{sec:gpgpu}
In a recent essay, Herb Sutter, a leading authority on software development, reviews the current state of the software and computer industry \cite{HerbSutter}. The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}. The two essays spins around the fact that chip manufactures hit a frequency-wall in the beginning of the 21st century. Until then, all mainstream computer programs where typically running in a single thread on a single core, and a programmer could expect the software to annually gain performance without touching the code. This was the "free-lunch" era as depicted in Fig.\,\ref{fig:jungle}. The increase in processing power came mostly from an ever-growing clock frequency. However, power consumption and heat generation where also growing at the same rate, until the level of cooling required finally became to much in 2004.  From that point, chip manufactures had to pursue a different approach  than just increasing the clock frequency. The immediate solution was to embed several cores in one CPU chip. This insured continuation of Moore's law, since the number of transistors per chip could continue to grow. For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to continue to gain performance. It was a concurrency revolution, and  the free lunch was over.

@@

\begin{figure}
\centering
\subfigure[Threads, blocks and grid]{
	\includegraphics[width=0.4\textwidth]{img/cuda_threads.png}
}
\subfigure[Kernel]{
	%\includegraphics[width=0.5\textwidth]{img/kernel.pdf}
}
\caption{Depicts how GPU threads are grouped into blocks and arranged in a grid. One thread runs a copy of a kernel function, in this example an element-wise matrix square operation.}
\label{fig:gpu_grid}
\end{figure}

### 2014-02-08 ThesisJP/Introduction/introduction.tex
Started adding references.

\subsection{Paper II}
\textbf{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}\\
\textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C\:Nilsen, A.\:Austeng and S.\:Holm\\
{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. \todo{add pages}, Jan. 2014.}\\\\
The second paper aims at implementing real-time Capon beamforming for cardiac ultrasound imaging. Achieving this will facilitate further study of this method's \textit{in vivo} performance. This is something which has been sought for the past decade \todo{Add citations}. In \textbf{Paper I} arrays with no more than 32 elements were investigated. A linear array for cardiac ultrasound imaging typically has 64 or more elements. In \textbf{Paper VI}, which is summarized in \textbf{Paper II}, it is shown that our implementation from  \textbf{Paper I} does not reach the level of throughput required for real-time Capon beamforming of cardiac ultrasound imaging. The matrices that have to be inverted become to large, and the frames that need to be processed per second are too many.

@@

\subsection{Paper IV}
\textbf{Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}\\
\textbf{J.\:P.\:\AA{}sen}, E.\:Steen, G.\:Kiss, A.\:Thorstensen and S.\:I.\:Rabben\\
{\it Proc. SPIE Medical Imaging 2012, \todo{add pages}.}\\\\
The fourth paper propose an adaptive volume rendering method based on statistics derived from the blood pool of the left ventricle. Noise located in the blood pool tends to occlude cardiac tissue when rendered in 3D, and is often impossible to remove by adjusting global rendering parameters. This because ultrasound signals have high variability within blood as well as high variability within tissue, even when heavily smoothed. Delineation of the blood pool is done by a state estimation algorithm, capable of tracking the endocardium in real time. The final visualizations are compared, with respect to location of tissue, by using a state-of-the-art endocardium segmentation tool. The paper presents both quantitative and qualitative results supporting the method's improved capabilities of reducing tissue dropouts and spurious structures inside the blood pool, leading to a maximized amount of visible endocardium tissue. The proposed method is implemented on a GPU with interactive frame rates as the result. \todo{litt tafatt. m\aa sm\oe{}re p\aa litt}

\subsection{Paper V}
\textbf{Huygens on Speed: Interactive Simulation of Ultrasound Pressure Fields}\\
\textbf{J.\:P.\:\AA{}sen} and S.\:Holm\\
{\it Proc. IEEE Ultrasonics Symposium 2012 \todo{add pages}.}\\\\
The fifth paper presents an interactive simulations tool capable of simulating dense pressure fields from ultrasound arrays in real time. Simulation tools are heavily used by researchers in order to test out new algorithms and to deduce the performance of new array designs prior to manufacturing. However, dense simulations involve extensive calculation and computational time. With the presented simulator we therefore try to reduce the time for calculating pressure fields by means of GPU acceleration. The tool is based on Huygens' principle which describes diffraction caused by a slit as the superposition of several point sources located inside the opening. Thus, the simulator works by accumulating the contribution from a collection of point sources in a set of observation points. How the source and observation points are laid out is left for the user to decide. Point sources positioned along a line will for instance simulate an ultrasound array. The simulator is linked with both a Paint-like interface for interactive drawing of arrays, and a Matlab interface for precise scripting of array configurations. The main contributions in this paper is the increased performance of the GPU implementation compared with a CPU and Matlab\footnote{The Ultrasim toolbox \todo{add citation}} version. Second, the paint-like interface provides a neat way of demonstrating array beamforming principle in real time. We believe this to be of great value both for array designers and teachers.

\section{Main contributions}
The main contributions of this thesis are:
\begin{enumerate}
\item a GPU implementation of the Capon beamformer.
\item a GPU implementation of the beamspace Capon beamformer.
\item that real-time beamspace Capon beamforming is achieved for cardiac ultrasound imaging.
\item the first investigation of Capon beamforming applied on multiple frames in medical ultrasound imaging (simulated, \textit{in vitro}, and \textit{in vivo}).
\item an investigation of shift-invariance when the Capon beamformer is applied on consecutive frames.
\item a method for improved shift-invariance of the Capon beamformer.
\item a method for reduced blood-pool noise in volume renderings of cardiac ultrasound volumes.
\item a GPU implementation of this adaptive volume rendering method.
\item a GPU implementation of simple paint-like simulation tool for rapid visualization ultrasound pressure fields.
\end{enumerate}
In addition, the thesis provides several discussions on how to utilize the GPU to accelerate advanced ultrasound imaging algorithms.

@@

In \textbf{Paper II} we focused on achieving real time Capon processing for cardiac ultrasound by implementing the BS-Capon beamformer on a GPU. However, we also present the first medical ultrasound loops processed with ES- and BS-Capon beamforming.  The paper does not provide any detailed evaluation of the method's \textit{in vivo} performance. Yet, initially it has been proven hard to transfer all results obtained in simulations to \textit{in vivo} images. The improvements seen are minimal, especially with the number of calculations in mind. The beamformers are clearly able to decrees the lateral width of lateral point scatterers and thin structures \textit{in vivo}, and to sharpen step edges. On the other hand, the overall contrast is visually not improved and sometimes it is worse (Fig.\,6b of \textbf{Paper II}). Since the Capon beamformer only affect directional noise, it is clear that contrast will get worse in low \nom{SNR}{Signal-to-noise ratio} situations when signal cancellation occurs. The oversampling approach presented in \textbf{Paper III} might improve on this, but it will only reduce signal cancellation if the phase shifts across the aperture are close to linear and small. Another issue is that bright points with a wide lateral profile exhibits better visual contrast when surrounded by speckle noise than points with narrow lateral profiles. In future work a detailed investigation of why some results, especially the contrast, obtained in simulations are not transferable to \textit{in vivo} needs to be performed. It will also be crucial to show larger improvements \textit{in vivio} than what is presented in \textbf{Paper II}. Both to justify all our computations, and to show an image which has both highly improved resolution and contrast. As pointed out at the end of \textbf{Paper II}, cardiac ultrasound imaging might not be the best modality for Capon beamforming. Cardiac ultrasound imaging was selected because it is the medical ultrasound modality which requires most calculations per second. An application where the focus is detection of closely spaced point targets, might be better suited. A better suited adaptive beamformer for cardiac applications might be the low-complexity adaptive beamformer (\nom{LCA beamformer}{Low complexity adaptive beamformer}), where the Capon optimization problem is applied on a set of predefined windows. This method has linear complexity and shares many properties with the Capon beamformer. If only minor improvements are obtained, it would be easier to ignore this fact if the algorithm is less costly. Finally, it would also be interesting to see how the Eigenspace-based Capon beamformer performs on cardiac images, and how its absence of a distortionless criterion will influence the image spatially and temporally.

@@

In \textbf{Paper3} we investigate the inherent self nulling involved with Capon beamforming, and how this effect comes naturally into play when the beamformer is subjected to linear steering vector errors caused by lateral undersampling. These errors are naturally reduced when the lateral sampling is increased. This simple observation is something which has been missing in the literature on Capon beamforming for medical ultrasound imaging. The reason could be ignorance or its possible negative impact on imaging frame rate. A lack of studies where Capon beamforming is performed on consecutive frames might also explain its absence. Fortunately, as shown in \textbf{Paper III}, it is possible to maintain the same frame rate as with DAS beamforming by applying oversampling on receive using a set of steering vectors (phase delays). It also turns out that more matrix inversions are not needed either. Oversampling on receive will, however,  increase the penalty of applying post-processing, like filtering,  on scan-grid data. %The time taken to convert data from a polar scan grid to a Cartesian display grid will also increase. The Cartesian grid should also have high resolution in order to preserve the un-sharpening introduced by Capon beamforming. 

@@

With improved shift-invariance it will also be interesting to see if high resolution beamforming could act as a post-processing step for other algorithms. For example to increase the precision of speckle and edge tracking. In the case where the image is generated for post processing it does not need to be visually pleasant anymore, as long as signal cancellation is avoided. Good looking speckle, or DAS-like speckle, has been the driving force behind many of the robustification techniques and the default parameters seen in the literature. Hence, more aggressive parameters might be used if the image produced is intended as input to another algorithm.

@@

One limitation of the algorithm described in \textbf{Paper IV} is its dependence on a successful tracking. If the tracking fails the proposed method will fail.  Another issue is the view-dependent visibility, where an object could become visible or disappear while rotating or if the local statistics change a lot during the cardiac cycle. This means that the rendering will be come non-intuitive to interact with and unrealistic to watch. However, the method could still be used to clean up still images of standard cardiac views. The paper does also propose an interesting method for automatic tuning of rendering parameters. Figure 3d in \textbf{Paper IV} actually shows that an optimal threshold could be found based on global and not only local statistics. Even though the visualization based on global statistics has more errors, with respect to endocardium visibility, the tuning of the opacity function is still fully automatic if an iterative search for the minima in Figure 3d is performed. The rendering will then be realistic, but still dependent on a successful tracking for the auto-adjustment to work. An automatically tuned opacity function based on global statistics combined with an improved smoothing scheme could be the future for cardiac ultrasound imaging. It would also be interesting to see if statistics based on the window selected by the Capon or the LCA beamformer could be utilized to adjust the opacity. It should also be possible to use this statistic as input to image filtering. An adaptively selected window will be symmetric in homogeneous regions and left-right shifted close to tissue interfaces. In that way we can apply less filtering and higher opacity on edges. If this will be different than just looking at neighboring pixels is to be seen.

@@

The fifth paper stands out from the other papers by not dealing with an adaptive processing technique. It is still, however, centered around the topic of implementing algorithms for ultrasound imaging on GPUs. Ultrasound simulations are typically used for research and sometimes also to pre-calculate configurations of a given scan sequence in order to achieve a certain pressure in a given area. The continuously growth in computational power will, in the future, make it possible to evaluate a scan sequence on the fly, removing the need for lookup tables. Estimation of the mechanical and thermal index could also be improved. Achieving higher processing speed for an excising algorithm by implementing it on a new architecture is not academically relevant in its own. It has to combined with either a good analysis of what special adjustments that had to be made to the algorithm in order for the increase to happen or a suggestion of how to utilize the speed gained. These are points we have tried to answer in both \textbf{Paper I}, \textbf{II},  \textbf{III}, and \textbf{V}.  In \textbf{Paper V} the increased speed makes an interactive paint program feasible. As of January 2014 the simulator has been download 200 times and counting. It also has a "whopping" 1000 views on youtube. 

>>>>

\subsection{Paper II}
\textbf{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}\\
\textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C\:Nilsen, A.\:Austeng and S.\:Holm\\
{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. \todo{add pages}, Jan. 2014.}\\\\
The second paper aims at implementing real-time Capon beamforming for cardiac ultrasound imaging. Achieving this will facilitate further study of the method's \textit{in vivo} performance. This is something which has been stated as further work in many publications on Capon beamforming for medical ultrasound imaging the past decade. In \textbf{Paper I} arrays with no more than 32 elements were investigated. A linear array for cardiac ultrasound imaging typically has 64 or more elements. In \textbf{Paper VI}, which is summarized in \textbf{Paper II}, it is shown that our implementation from  \textbf{Paper I} does not reach the level of throughput required for real-time Capon beamforming of cardiac ultrasound imaging. The matrices that have to be inverted become to large, and the frames that need to be processed per second are too many.

@@

\subsection{Paper IV}
\textbf{Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}\\
\textbf{J.\:P.\:\AA{}sen}, E.\:Steen, G.\:Kiss, A.\:Thorstensen and S.\:I.\:Rabben\\
{\it Proc. SPIE Medical Imaging 2012, \todo{add pages}.}\\\\
The fourth paper propose an adaptive volume rendering method based on statistics derived from the blood pool of the left ventricle. Noise located in the blood pool tends to occlude cardiac tissue when rendered in 3D, and is often impossible to remove by adjusting global rendering parameters. This because ultrasound signals have high variability within blood as well as high variability within tissue, even when heavily smoothed. Delineation of the blood pool is done by a state estimation algorithm, capable of tracking the endocardium in real time. The final visualizations are compared, with respect to location of tissue, by using a state-of-the-art endocardium segmentation tool. The paper presents both quantitative and qualitative results supporting the method's improved capabilities of reducing tissue dropouts and spurious structures inside the blood pool, leading to a maximized amount of visible endocardium tissue. The proposed method is implemented on a GPU with real-time frame rates as the result. This results shows that with modern GPUs it is possible to add more advanced visualization to an ultrasound imaging system and still have real-time performance.

\subsection{Paper V}
\textbf{Huygens on Speed: Interactive Simulation of Ultrasound Pressure Fields}\\
\textbf{J.\:P.\:\AA{}sen} and S.\:Holm\\
{\it Proc. IEEE Ultrasonics Symposium 2012 \todo{add pages}.}\\\\
The fifth paper presents an interactive simulations tool capable of simulating dense pressure fields from ultrasound arrays in real time. Simulation tools are heavily used by researchers in order to test out new algorithms and to deduce the performance of new array designs prior to manufacturing. However, dense simulations involve extensive calculation and computational time. With the presented simulator we aim at reducing the time needed for calculating pressure fields by means of GPU acceleration. The tool is based on Huygens' principle which describes diffraction caused by a slit as the superposition of several point sources located inside the opening. Thus, the simulator works by accumulating the contribution from a collection of point sources in a set of observation points. How the source and observation points are laid out is left for the user to decide. Point sources positioned along a line will for instance simulate an ultrasound array. The simulator is linked with both a Paint-like interface for interactive drawing of arrays, and a Matlab interface for precise scripting of array configurations. The main contributions in this paper is the increased performance of the GPU implementation compared with a CPU and Matlab\footnote{The Ultrasim toolbox \todo{add citation}} version. Second, the paint-like interface provides a neat way of demonstrating array beamforming principle in real time. We believe this to be of great value both for array designers and teachers.

\section{Main contributions}
The main contributions of this thesis are:
\begin{enumerate}
\item a GPU implementation of the Capon beamformer.
\item a GPU implementation of the beamspace Capon beamformer.
\item that real-time beamspace Capon beamforming is achieved for cardiac ultrasound imaging.
\item the first investigation of Capon beamforming applied on multiple frames in medical ultrasound imaging (simulated, \textit{in vitro}, and \textit{in vivo}).
\item an investigation of shift-invariance when the Capon beamformer is applied on consecutive frames.
\item a method for improved shift-invariance of the Capon beamformer.
\item a method for reduced blood-pool noise in volume renderings of cardiac ultrasound volumes.
\item a GPU implementation of this adaptive volume rendering method.
\item a GPU implementation of simple Paint-like simulation tool for rapid visualization ultrasound pressure fields.
\end{enumerate}
In addition, the thesis provides several discussions on how to utilize the GPU to accelerate advanced ultrasound imaging algorithms.

@@

In \textbf{Paper II} we focused on achieving real time Capon processing for cardiac ultrasound by implementing the BS-Capon beamformer on a GPU. However, we also present the first medical ultrasound loops processed with ES- and BS-Capon beamforming.  The paper does not provide any detailed evaluation of the method's \textit{in vivo} performance. Yet, initially it has been proven hard to transfer all results obtained in simulations to \textit{in vivo} images. The improvements seen are minimal, especially with the number of calculations in mind. The beamformers are clearly able to decrees the lateral width of lateral point scatterers and thin structures \textit{in vivo}, and to sharpen step edges. On the other hand, the overall contrast is visually not improved and sometimes it is worse (Fig.\,6b of \textbf{Paper II}). Since the Capon beamformer only affect directional noise, it is clear that contrast will get worse in low \nom{SNR}{Signal-to-noise ratio} situations when signal cancellation occurs. The oversampling approach presented in \textbf{Paper III} might improve on this, but it will only reduce signal cancellation if the phase shifts across the aperture are close to linear and small. Another issue is that bright points with a wide lateral profile exhibits better visual contrast when surrounded by speckle noise than points with narrow lateral profiles. In future work a detailed investigation of why some results, especially the contrast, obtained in simulations are not transferable to \textit{in vivo} needs to be performed. It will also be crucial to show larger improvements \textit{in vivio} than what is presented in \textbf{Paper II}. Both to justify all our computations, and to show an image which has both highly improved resolution and contrast. As pointed out at the end of \textbf{Paper II}, cardiac ultrasound imaging might not be the best modality for Capon beamforming. Cardiac ultrasound imaging was selected because it is the medical ultrasound modality which requires most calculations per second. An application where the focus is detection of closely spaced point targets, might be better suited. A better suited adaptive beamformer for cardiac applications might be the low-complexity adaptive beamformer (\nom{LCA beamformer}{Low complexity adaptive beamformer}), where the Capon optimization problem is applied on a set of predefined windows. This method has linear complexity and shares many properties with the Capon beamformer. If only minor improvements are obtained, it would be easier to ignore this fact if the algorithm is less computationally costly. Finally, it would also be interesting to see how the Eigenspace-based Capon beamformer performs on cardiac images, and how its absence of a distortionless criterion will influence the image spatially and temporally.

@@

In \textbf{Paper3} we investigate the inherent self nulling involved with Capon beamforming, and how this effect comes naturally into play when the beamformer is subjected to linear steering vector errors caused by lateral undersampling. These errors are naturally reduced when the lateral sampling is increased. This simple observation is something which has been missing in the literature on Capon beamforming for medical ultrasound imaging. The reason could be ignorance or its possible negative impact on imaging frame rate. A lack of studies where Capon beamforming is performed on consecutive frames might also explain the lack of comments. Fortunately, as shown in \textbf{Paper III}, it is possible to maintain the same frame rate as with DAS beamforming by applying oversampling on receive using a set of steering vectors (phase delays). It also turns out that more matrix inversions are not needed either. Oversampling on receive will, however,  increase the penalty of applying post-processing, like filtering,  on scan-grid data. %The time taken to convert data from a polar scan grid to a Cartesian display grid will also increase. The Cartesian grid should also have high resolution in order to preserve the un-sharpening introduced by Capon beamforming. 

@@

With improved shift-invariance it will also be interesting to see if high resolution beamforming could act as a post-processing step for other algorithms. For example to increase the precision of speckle and edge tracking. In the case where the image is generated for post processing it does not need to be visually pleasant anymore, as long as signal cancellation is avoided. Good looking speckle, or DAS-like speckle, has been the driving force behind many of the robustification techniques and the default parameters seen in the literature. Hence, more aggressive parameters might be used if the image produced is intended as input to another algorithm and not for the human eye.

@@

One limitation of the algorithm described in \textbf{Paper IV} is its dependence on a successful tracking. If the tracking fails the proposed method will fail.  Another issue is the view-dependent visibility, where an object could become visible or disappear while rotating or if the local statistics change a lot during the cardiac cycle. Even if the method successfully removes noise, this means that the rendering will become non-intuitive to interact with and unrealistic to watch. However, the method could still be used to clean up still images of standard cardiac views. The paper does also propose an interesting method for automatic tuning of rendering parameters. Figure 3d in \textbf{Paper IV} actually shows that an optimal threshold could be found based on global and not only local statistics. Even though the visualization based on global statistics has more errors, with respect to endocardium visibility, the tuning of the opacity function is still fully automatic if an iterative search for the minima in Figure 3d is performed. A global opacity transfer function will assure a rendering which are realistic, but it will still be dependent on a successful tracking for the auto-adjustment to work. An automatically tuned opacity function based on global statistics combined with an improved smoothing scheme could be the future for cardiac ultrasound imaging. It would also be interesting to see if statistics based on the window selected by the Capon or the LCA beamformer could be utilized to adjust the opacity. It should also be possible to use this statistic as input to image filtering. An adaptively selected window will be symmetric in homogeneous regions and left-right shifted close to tissue interfaces. In that way we can apply less filtering and higher opacity on edges. If this will be different than just looking at neighboring pixels is to be seen.

@@

The fifth paper stands out from the other papers by not dealing with an adaptive processing technique. It is still, however, centered around the topic of implementing algorithms for ultrasound imaging on GPUs. Ultrasound simulations are typically used for research and sometimes also to pre-calculate configurations of a given scan sequence in order to achieve a certain pressure in a given area. The continuously growth in computational power will, in the future, make it possible to evaluate a scan sequence on the fly, removing the need for lookup tables. Estimation of the mechanical and thermal index could also be improved. Achieving higher processing speed for an excising algorithm by implementing it on a new architecture is not academically relevant in its own. It has to combined with either a good analysis of what special adjustments that had to be made to the algorithm in order for the increase to happen or a suggestion of how to utilize the speed gained. These are points we have tried to answer in both \textbf{Paper I}, \textbf{II},  \textbf{IV}, and \textbf{V}.  In \textbf{Paper V} the increased speed makes an interactive paint program feasible. As of January 2014 the simulator has been download 200 times and counting. It also has a "whopping" 1000 views on youtube. 
